Concurrent Programming with CXL

### 2020-05-18 book/paper.tex
Starting to add Peterson's proof
got rid of "," part of assert
Finished draft of Peterson chapter

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \textbf{while} statements and \textbf{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not support floating point or lists natively;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

\chapter{The CXL Machine}

@@

>>>>

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not support floating point or lists natively;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

\chapter{The CXL Machine}
\label{ch:cxlmachine}

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is no where near the critical section
or has given way.
Running the algorithm though CXL shows that it satisfies both mutual
exclusion and progress.

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
at any point of time the CXL machine is in a specific state.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process is
updated).  We call that a \emph{step}.  Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace} is a
sequence of steps starting from the initial state.  When making a
step, there are two sources of non-determinism in CXL.  One is when
in some state more than one process can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.

A good way to prove mutual exclusion is through induction on
the number of steps.
In such a proof, we need a so-called
\emph{inductive invariant} $I$, which is some predicate that holds
over the state.  We need $I$ to satisfy the following:
\begin{itemize}
\item (ME1) $I$ should imply mutual exclusion.
\item (ME2) $I$ should hold in the initial state.
\item (ME3) If $I$ holds over some state $s$, and some process in the
state takes a step, then $I$ should also hold in the resulting state.
\end{itemize}

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly satisfies ME1 and ME2.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every \emph{reachable state}.
Unfortunately, it is not an \emph{inductive} invariant and therefore does
not satisfy ME3.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

\begin{figure}
\begin{code}
\verbatiminput{../PetersonProof.cxl}
\end{code}
\caption{Peterson's Algorithm with Inductive Invariant}
\label{fig:petersonproof}
\end{figure}

Let $C(\mathtt{self}) = \mathtt{flags}[1 - \mathtt{self}] \land
\mathtt{turn} = 1 - \mathtt{self}$, that is, the condition on the \texttt{while} loop.
For Peterson's Algorithm, an inductive invariant that works well is:
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
Let's call it $I_p(i)$.

Figure~\ref{fig:petersonproof} formalizes $I_p(i)$ in CXL.
The figure uses the CXL \texttt{atomic} statement for the first time.
For now, all you need to know is that it turns the embedded statement into
a single atomic step.  The label \texttt{@gate} thus refers exactly to
this statement.  You can run Figure~\ref{fig:petersonproof} to determine
that $I_p(i)$ is indeed an invariant for $i = 0, 1$.

To see that $I_p(i)$ implies mutual exclusion, suppose not.  Then
when both process 0 and process 1 are in the critical section, the
following must hold:
$(\lnot C(0) \lor \mathtt{process}(1)@gate) \land
 (\lnot C(1) \lor \mathtt{process}(0)@gate)$.
We know that both flags are set.
We also know that neither process 1 nor process 2 is at label \texttt{@gate}
(because they are both at label \texttt{@cs}),
so this simplifies to $(\lnot C(0)) \land (\lnot C(1))$.
So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
logical impossibility.  Thus $I_p(i)$ satisfies ME1.

$I_p(i)$ satisfies ME2 trivially, because in the initial state no process
is in the critical section.

Finally we have to show that $I_p(i)$ satisfies ME3. Without loss of
generality, suppose $i=0$ (a benefit from the fact that the algorithm is
symmetric for both processes).  We have to show that if we are in a state
in which $I_p(0)$ holds, then any step will result in a state in which
$I_p(0)$ continues to hold.
Since $I_p(0)$ holds, process 0 is at label \texttt{@cs}.  If process 0
were to take a step, then in the next state process 0 would be no longer
at that label and $I_p(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by process 1.

From $I_p(0)$ we know that one of the following cases must hold before
process 1 takes a step:
\begin{itemize}
\item \texttt{flags[1] = False};
\item \texttt{turn = 0};
\item process 1 is at label \texttt{@gate}.
\end{itemize}

Let us consider each of these cases.
In the first case, if process 1 takes a step, there are two possibilities:
either $flags[1]$ will still be \texttt{False} (in which case the first case
continues to hold), or $flags[1]$ will be \texttt{True}
and process 1 will be at label \texttt{@gate} (in which case the third case
will hold).
We know that process 1 never sets \texttt{turn} to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if process 1 is at label \texttt{@gate} before the step, then after
the step \texttt{turn} will equal 0, and therefore the second case will hold
after the step.

We have now demonstrated mutual exclusion in Peterson's Algorithm in two
different ways: one by letting CXL explore all possible executions, the
other using an inductive invariant and proof by induction.  The former
is certainly easier, but it does not provide intuition for why the
algorithm works.  The second provides much more insight.  We therefore
encourage to include inductive invariants in your CXL code.

A cool anecdote is the following.  When the author of CXL had to teach
Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
page.  The page claimed that the following predicate was invariant:
if process $i$ is in the critical section, then $\lnot C(i)$ (i.e.,
$I_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
To demonstrate that this predicate is not invariant, you can remove the
disjunct from Figure~\ref{fig:petersonproof} and run it to get a
counterexample.

This anecdote suggests the following.  If you need to do an induction
proof of an algorithm, you have to come up with an inductive invariant.
Before trying to prove the algorithm, you can check that the predicate is
at least invariant by testing it using CXL.  Doing so could potentially
avoid wasting your time on a proof that will not work because the
predicate is not invariant, and therefore not an inductive invariant either.

To finish the story, the author of CXL fixed the Wikipedia page, but there
is another cool story.  Years later a colleague of the author, teaching
the same course, asked if the first two assignments (setting \texttt{flags[self]})
to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
After all, they are different variables assigned independent values---in a
sequential program one could surely swap the two assignments.

\begin{quote}
What do you think?
\begin{itemize}
\item See if you can figure out for yourself if the two assignments can be
reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
the two assignments and see what happens.
\end{itemize}
\end{quote}
